%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%     File: ExtendedAbstract_abstr.tex                               %
%     Tex Master: ExtendedAbstract.tex                               %
%                                                                    %
%     Author: Andre Calado Marta                                     %
%     Last modified : 2 Dez 2011                                     %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The abstract of should have less than 500 words.
% The keywords should be typed here (three to five keywords).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Abstract
%%
\vspace{-0.42cm}

\begin{abstract}

The liberalization of the air transport market in Europe led to unprecedented air traffic growth and the
emergence of Low-Cost Carriers (LCCs), which transformed the aviation industry. This
evolution intensified the need for airports to operate efficiently and adapt continuously to changing
market conditions. Additionally, the growing adoption of concession models in airport management
emphasizes the importance of continuous performance evaluation, to provide objective measures of
efficiency to public regulators.
The present dissertation evaluates the performance of 41 Iberian airports from 2016 to 2023 by
combining Data Envelopment Analysis (DEA) with a two-stage truncated regression and applying a bootstrap
procedure for bias correction. Additionally, the Malmquist Productivity Index (MPI) is used to assess
productivity changes over time.
The results reveal a low average efficiency across airports of the Iberian Peninsula (0.522, excluding
COVID-19 years), especially in the Spanish network, and identify the best and worst performing airports.
The truncated regression highlights the importance of rail intermodal terminals on airports, and indicates
that airports located on Islands, those with Military operations, or those with a higher cargo orientation
tend to be more efficient. The MPI allowed to identify the effects of the COVID-19 pandemic on each
component of airport efficiency, while also presented the airports with best and worst efficiency evolution
in the studied period.
\\
%%
%% Keywords (max 5)
%%
\noindent{{\bf Keywords:}} DEA, Truncated Regressions, MPI, Airport Efficiency, Portugal and Spanish Airports   \\

\end{abstract}

